---
output:
  github_document:
    html_preview: false
---

# zigdist

`zigdist` is an experimental demonstration that integrates the [Zig programming language](https://ziglang.org/) into an R package. It provides high-performance distance metric calculations as a proof-of-concept. While Zig is not currently supported on CRAN, its exceptional performance, simple FFI, data-oriented design, and robust systems programming capabilities suggest it could be of significant interest to the R package development community.

## Distance methods

- Euclidean (`zd_euclidean()`): Pairwise row distance and cross-distance for numeric matrices, supporting missing values (partial distance scaling)
- Gower (`zd_gower()`): Mixed-type distance computations (scaling numeric features by range, matching categorical features), supporting missing values (pairwise deletion)

These distance methods use multi-threaded parallel execution by specifying `num_threads > 1`.

## Installation

You can install the development version of `zigdist` from GitHub:

```r
# install.packages("remotes")
remotes::install_github("brownag/zigdist")
```

Note: A Zig compiler (v0.13.0 or later) must be installed on your system and available in your `PATH` to compile the package from source.

## Quick Start

```{r example}
library(zigdist)

# Euclidean
x <- matrix(rnorm(15), nrow = 5, ncol = 3)
zd_euclidean(x)

# Gower (supports mixed types: numeric, nominal factors, and ordered factors)
df <- data.frame(
  a = c(1, 2, 5),
  b = factor(c("A", "B", "A")),
  c = ordered(c("Low", "Medium", "High"), levels = c("Low", "Medium", "High"))
)
zd_gower(df)
```

## Correctness and Equivalence

To demonstrate functional equivalence, we verify that the outputs of `zigdist` match the standard R implementations (`stats::dist` and `cluster::daisy`) using `all.equal()`:

```{r verification}
# Euclidean vs. stats::dist
x_check <- matrix(rnorm(300), nrow = 20, ncol = 15)
all.equal(zd_euclidean(x_check), as.matrix(dist(x_check)), check.attributes = FALSE)

# Gower vs. cluster::daisy
library(cluster)
df_check <- data.frame(
  n = rnorm(20),
  f = factor(sample(letters[1:3], 20, replace = TRUE)),
  c = factor(sample(c("Yes", "No"), 20, replace = TRUE))
)
all.equal(zd_gower(df_check), as.matrix(daisy(df_check, metric = "gower")), check.attributes = FALSE)
```

## Performance Benchmarks

Below are benchmarks comparing `zigdist` against standard R implementations, using the `microbenchmark` package.

```{r benchmark_setup, message=FALSE, warning=FALSE}
library(microbenchmark)
library(cluster)
```

### Euclidean

We compare `zigdist::zd_euclidean` (single-threaded and multi-threaded) against R's built-in `stats::dist` (converted to a full matrix) on a $500 \times 100$ numeric matrix.

```{r benchmark_euclidean}
set.seed(42)
x_bench <- matrix(rnorm(50000), nrow = 500, ncol = 100)

bench_eucl <- microbenchmark(
  `zigdist 1 thread` = zd_euclidean(x_bench, num_threads = 1L),
  `zigdist 4 thread` = zd_euclidean(x_bench, num_threads = 4L),
  stats_dist = as.matrix(dist(x_bench)),
  times = 50
)
print(bench_eucl)
```

```{r calc_eucl_speedup, echo=FALSE}
sum_eucl <- summary(bench_eucl)
eucl_speedup_1t <- sum_eucl$median[sum_eucl$expr == "stats_dist"] / sum_eucl$median[sum_eucl$expr == "zigdist 1 thread"]
eucl_speedup_4t <- sum_eucl$median[sum_eucl$expr == "stats_dist"] / sum_eucl$median[sum_eucl$expr == "zigdist 4 thread"]
```

Even on a single thread (`num_threads = 1L`), `zigdist` significantly outperforms R's native `stats::dist` (written in optimized C) by more than **`r round(eucl_speedup_1t, 1)`x**. Spawning multiple threads (e.g., `num_threads = 4L`) achieves a **`r round(eucl_speedup_4t, 1)`x speedup**.

### Gower

We compare `zigdist::zd_gower` (single-threaded and multi-threaded) against `cluster::daisy(..., metric = "gower")` (converted to a full matrix) on a mixed-type data frame with 500 rows and 6 columns (3 numeric, 3 categorical).

```{r benchmark_gower}
set.seed(42)
N_bench <- 500
df_bench <- data.frame(
  n1 = rnorm(N_bench),
  n2 = runif(N_bench),
  n3 = rnorm(N_bench, mean = 10, sd = 2),
  c1 = factor(sample(letters[1:4], N_bench, replace = TRUE)),
  c2 = factor(sample(c("Yes", "No"), N_bench, replace = TRUE)),
  c3 = factor(sample(colors()[1:10], N_bench, replace = TRUE))
)

bench_gow <- microbenchmark(
  `zigdist 1 thread` = zd_gower(df_bench, num_threads = 1L),
  `zigdist 4 thread` = zd_gower(df_bench, num_threads = 4L),
  cluster_daisy = as.matrix(daisy(df_bench, metric = "gower")),
  times = 20
)
print(bench_gow)
```

```{r calc_gow_speedup, echo=FALSE}
sum_gow <- summary(bench_gow)
gow_speedup_1t <- sum_gow$median[sum_gow$expr == "cluster_daisy"] / sum_gow$median[sum_gow$expr == "zigdist 1 thread"]
```

For mixed-type data, `zigdist` represents a substantial improvement over `cluster::daisy`, calculating distances **`r round(gow_speedup_1t, 1)`x faster** in single-threaded mode. For smaller calculations, running sequentially on a single thread is recommended to eliminate thread-spawning overhead.
