---
title: "Introduction to zigdist"
date: "`r Sys.Date()`"
output: litedown::html_format
vignette: >
  %\VignetteIndexEntry{Introduction to zigdist}
  %\VignetteEngine{litedown::vignette}
  %\VignetteEncoding{UTF-8}
---

# Introduction to zigdist

The `zigdist` package provides high-performance, minimal-dependency implementations of distance metrics in the Zig programming language. It currently implements:

- **Euclidean Distance** (self-distance and cross-distance for numeric matrices/data frames).
- **Gower (1971) Distance** (self-distance and cross-distance for mixed-type data frames containing numeric, categorical, and ordinal features).

## Euclidean Distance

To calculate the pairwise Euclidean distance matrix of a numeric matrix `x`:

```{r}
library(zigdist)

# Generate a random matrix
set.seed(42)
x <- matrix(rnorm(15), nrow = 5, ncol = 3)

# Calculate pairwise distances (returns 5x5 matrix)
zd_euclidean(x)
```

To calculate the cross-distance matrix between two matrices `x` (size $N \times P$) and `y` (size $M \times P$):

```{r}
y <- matrix(rnorm(9), nrow = 3, ncol = 3)

# Calculate cross distance (returns 5x3 matrix)
zd_euclidean(x, y)
```

## Gower Distance

Gower distance is designed for mixed data types. Numeric and ordinal columns are scaled by their range:

$$d_{ijk} = \frac{|x_{ik} - x_{jk}|}{\text{range}(x_{\cdot k})}$$

Nominal (categorical/logical) columns are compared using simple matching:

$$d_{ijk} = I(x_{ik} \neq x_{jk})$$

The total Gower distance is the weighted average of these components:

$$D_{ij} = \frac{\sum_k w_k d_{ijk}}{\sum_k w_k}$$

### Ordinal Factors (Ordered Categories)
Ordinal factors are natively supported. Following Podani's (1999) metric approach, `zd_gower` automatically converts ordered factors to their integer ranks ($1, 2, \dots$) and handles them as range-scaled numeric columns. This preserves the relative ordering and "closeness" of the categories (e.g., the distance between `Low` and `Medium` is smaller than between `Low` and `High`), matching R's standard `cluster::daisy` behavior.

### Example

```{r}
# Create a mixed-type data frame with numeric, nominal, and ordinal columns
df_x <- data.frame(
  a = c(1, 2, 5),
  b = factor(c("A", "B", "A")),
  c = ordered(c("Low", "Medium", "High"), levels = c("Low", "Medium", "High"))
)

# Calculate pairwise Gower distances
zd_gower(df_x)
```

You can also compute cross-distances between two mixed-type datasets, and provide custom column weights:

```{r}
df_y <- data.frame(
  a = c(2, 4),
  b = factor(c("A", "C")),
  c = ordered(c("Medium", "High"), levels = c("Low", "Medium", "High"))
)

# Weighted Gower cross-distance
zd_gower(df_x, df_y, weight = c(2.0, 0.5, 0.5))
```

Columns with zero range (i.e. constant values) are automatically ignored (excluded from the calculations) without raising division-by-zero errors.

## References

- Gower, J. C. (1971). A general coefficient of similarity and some of its properties. *Biometrics*, 27(4), 857–871. https://doi.org/10.2307/2528823
- Podani, J. (1999). Extending Gower's general coefficient of similarity to ordinal characters. *Taxon*, 48(2), 331–340. https://doi.org/10.2307/1224024

